---
title: "gam fit"
author: "JE Hug"
output: pdf_document
---

```{r setup, include=TRUE, message = FALSE}
library(tidyverse)
library(magrittr)
library(dplyr)
library(mgcv)
```


```{r Read in Data}
# read in df
huge <- read.csv(file = "../../Full_Data_Second_Clean/2020_08_Covariates_Pacific.csv")

# take away NA
covariatesDF <- na.omit(huge)

# select columns we need ( a single buoy)

all_buoys_no_date <- covariatesDF %>% dplyr::select(PLATFORM_NUMBER, LONGITUDE, LATITUDE, TEMP_ADJUSTED, PRES_ADJUSTED )


all_buoys_no_date %<>% dplyr::filter(PRES_ADJUSTED > 15)


```



```{r}
library(mgcv)

# pick a left out buoy to see what the predicted buoy looks like

leftout <- all_buoys_no_date %>% dplyr::filter(PLATFORM_NUMBER == 2902748 & LONGITUDE == 151.87 & LATITUDE == 32.502) %>% 
  select(-TEMP_ADJUSTED)

# take the data except the left out buoy

test_buoy <- all_buoys_no_date %>% dplyr::filter(PLATFORM_NUMBER != 2902748 & LONGITUDE != 151.87 & LATITUDE != 32.502)


# fit the model
# we use an interaction of longitude and latitude
fit <- gam(formula = TEMP_ADJUSTED ~ s(LONGITUDE,LATITUDE) + s(PRES_ADJUSTED) , data=test_buoy, method = "REML")

# summary of the fit
summary(fit)



# predict the leftout temperatures from the GAM

TEMP_ADJUSTED <- predict.gam(object = fit,newdata = leftout )

# bind 

dearlord <- cbind(leftout,TEMP_ADJUSTED)

# plot the predicted

dearlord %>% ggplot(aes(x=TEMP_ADJUSTED,y=PRES_ADJUSTED))+
  geom_point()

# plot the true

all_buoys_no_date %>% filter(PLATFORM_NUMBER == 2902748 & LONGITUDE == 151.87 & LATITUDE == 32.502) %>% ggplot(aes(x=TEMP_ADJUSTED,y=PRES_ADJUSTED))+
  geom_point()
```


Some thoughts on this, we don't capture the kink but the overall shape may be okay?
We can try and bin blocks in the region and fit a gamm (mixed model) where 
our random whatevers (idk smooths?) are over these blocks. Alternatively 
I used the interaction here but I'm not sure what it does I went from this 
GAM guide https://noamross.github.io/gams-in-r-course/chapter3 , also see 
Simon Wood's text Generalized additive models, from which the mgcv package 
comes from! A GP will be very difficult to do, but we can try but again I'm concerned 
about capturing the kink.




# correlating surface temp to typhoons


```{r}

library(gamm4)
binned <- test_buoy %>% mutate(location_bin=case_when(
  LATITUDE > 25 & LATITUDE < 30 ~ "a",
  LATITUDE > 30 & LATITUDE < 35 ~ "b",
  LATITUDE > 35 & LATITUDE < 40 ~ "c",
  LATITUDE > 40 & LATITUDE < 45 ~ "d",
  LATITUDE > 45 & LATITUDE < 50 ~ "e",
  LATITUDE > 50 & LATITUDE < 55 ~ "f"
))

binned$location_bin <- as.factor(binned$location_bin)




fit2 <- gamm4(formula = TEMP_ADJUSTED ~ s(LONGITUDE,LATITUDE) + s(PRES_ADJUSTED, by = location_bin) ,random=~(1|location_bin), data=binned)



```


```{r IMPORT TYPHOON DATA}

typh <- read_csv("../typhoon_data/bwp_2019/2019_typh_data.csv")

typh$yr_cyc_num <- gsub(pattern = ",",replacement = "",x = typh$yr_cyc_num)
typh$mslp <- as.numeric(gsub(pattern = ",",replacement = "",x = typh$mslp))

# NEED TO REPLACE with NEGATIVE SIGNS FOR W/S

typh$lat <- as.numeric(gsub(pattern = "(N,|S,)",replacement = "",x = typh$lat))/10
typh$long <- as.integer(gsub(pattern = "(E,|W,)",replacement = "",x = typh$long))/10


firstfive <- typh %>% group_by(yr_cyc_num) %>% filter(row_number()==c(1:5))
colnames(firstfive) <- c("yr_cyc_num","date_time","LATITUDE","LONGITUDE","vmax","mslp","typh_grade")
```


```{r}
withall <- cbind(firstfive, PRES_ADJUSTED = 10)
predictedST <- predict.gam(fit, withall)

withpred <- cbind(withall,surTemp = predictedST)

summary(withpred$surTemp)
```


# mapping of thermocline depth

```{r}

## Useful buoys -- Deep enough to use the variable representative isotherm 
## formula (> 400 meters) and should also have surface temperature values.
## 
## For the variable isotherm formula, the equation is as follows:
## 
##  TT = T(MLD) – 0.25[T(MLD) – T(400m)]
##
## TT -- Thermocline Temperature
## T(MLD) -- Temperature at the base of the mixed layer
## T(400m) -- Temperature at 400 meters
## 
## For T(MLD), we use surface temperature minus 0.8.
##
## See Fiedler 2010, pg. 319, for more details. Link below:
## https://aslopubs.onlinelibrary.wiley.com/doi/pdfdirect/10.4319/lom.2010.8.313
## 


deepBuoys <- all_buoys_no_date %>% filter(PRES_ADJUSTED > 400 & TEMP_ADJUSTED < 35) %>%
               group_by(PLATFORM_NUMBER , LONGITUDE , LATITUDE) %>%
               summarise( TEMP_DEEP = max(TEMP_ADJUSTED ) )

surfaceBuoys <- all_buoys_no_date %>% filter(PRES_ADJUSTED < 30 & PRES_ADJUSTED > 10) %>%
                group_by(PLATFORM_NUMBER , LONGITUDE , LATITUDE) %>%
                summarise( SURFACE_TEMP = mean(TEMP_ADJUSTED) )

usefulBuoys <- inner_join( deepBuoys , surfaceBuoys )

usefulBuoys$TEMP_MLD <- usefulBuoys$SURFACE_TEMP - 0.8

usefulBuoys$TEMP_THERMOCLINE <- usefulBuoys$TEMP_MLD - 0.25 * ( usefulBuoys$TEMP_MLD -
                                  usefulBuoys$TEMP_DEEP )

usefulBuoys$THERMOCLINE_DEPTH <- NA


## This for() loop adds the THERMOCLINE_DEPTH to each entry using the value
## immediately below the thermocline level.
##
## ~ 1-2% of the data is quite strange here: It's having trouble locating thermoclines
## for those particular buoys. Should investigate further.
##
## Also, this code is slow as f#@%, even though it runs.
for(i in 1:nrow(usefulBuoys)){
  single_working_buoy <- all_buoys_no_date %>%
                         filter( PLATFORM_NUMBER == usefulBuoys$PLATFORM_NUMBER[i] &
                                 LONGITUDE == usefulBuoys$LONGITUDE[i] &
                                 LATITUDE == usefulBuoys$LATITUDE[i] )
  
  dist_to_thermocline <- abs( single_working_buoy$TEMP_ADJUSTED - usefulBuoys$TEMP_THERMOCLINE[i] )
  temp_sorting_value <- sort( dist_to_thermocline )[3]
  
  rows_near_therm <- single_working_buoy[ dist_to_thermocline < temp_sorting_value , ]
  
  usefulBuoys$THERMOCLINE_DEPTH[i] <- min( rows_near_therm$PRES_ADJUSTED )
  
}


## Here is a group of weird buoys that don't fit or seem to have thermoclines. 
## They make up about ~1.5% of total data.
weirdBuoys <- usefulBuoys[ usefulBuoys$THERMOCLINE_DEPTH > 400 , ]

fitThermDepth <- gam(formula = THERMOCLINE_DEPTH ~ s(LONGITUDE,LATITUDE), data = usefulBuoys)
```


We will now import another months data and predict the thermocline depth there 
and see if the model does a good job.


```{r}
seperateMonth <- read.csv(file = "../../Full_Data_Second_Clean/2019_08_Covariates_Pacific.csv")


# take away NA
nona <- na.omit(seperateMonth)

# select columns we need ( a single buoy)

sepmonth <- nona %>% dplyr::select(PLATFORM_NUMBER, LONGITUDE, LATITUDE, TEMP_ADJUSTED, PRES_ADJUSTED )


sepmonth %<>% dplyr::filter(PRES_ADJUSTED > 15)

thermpred <- predict.gam(fitThermDepth, sepmonth)

sepmonthwp <- cbind(sepmonth,thermpred)



```

Get what we would have predicted had we the full buoy data

```{r}

## Useful buoys -- Deep enough to use the variable representative isotherm 
## formula (> 400 meters) and should also have surface temperature values.
## 
## For the variable isotherm formula, the equation is as follows:
## 
##  TT = T(MLD) – 0.25[T(MLD) – T(400m)]
##
## TT -- Thermocline Temperature
## T(MLD) -- Temperature at the base of the mixed layer
## T(400m) -- Temperature at 400 meters
## 
## For T(MLD), we use surface temperature minus 0.8.
##
## See Fiedler 2010, pg. 319, for more details. Link below:
## https://aslopubs.onlinelibrary.wiley.com/doi/pdfdirect/10.4319/lom.2010.8.313
## 


deepBuoys <- sepmonthwp %>% filter(PRES_ADJUSTED > 400 & TEMP_ADJUSTED < 35) %>%
               group_by(PLATFORM_NUMBER , LONGITUDE , LATITUDE) %>%
               summarise( TEMP_DEEP = max(TEMP_ADJUSTED ) )

surfaceBuoys <- sepmonthwp %>% filter(PRES_ADJUSTED < 30 & PRES_ADJUSTED > 10) %>%
                group_by(PLATFORM_NUMBER , LONGITUDE , LATITUDE) %>%
                summarise( SURFACE_TEMP = mean(TEMP_ADJUSTED) )

usefulBuoys <- inner_join( deepBuoys , surfaceBuoys )

usefulBuoys$TEMP_MLD <- usefulBuoys$SURFACE_TEMP - 0.8

usefulBuoys$TEMP_THERMOCLINE <- usefulBuoys$TEMP_MLD - 0.25 * ( usefulBuoys$TEMP_MLD -
                                  usefulBuoys$TEMP_DEEP )

usefulBuoys$THERMOCLINE_DEPTH <- NA


## This for() loop adds the THERMOCLINE_DEPTH to each entry using the value
## immediately below the thermocline level.
##
## ~ 1-2% of the data is quite strange here: It's having trouble locating thermoclines
## for those particular buoys. Should investigate further.
##
## Also, this code is slow as f#@%, even though it runs.
for(i in 1:nrow(usefulBuoys)){
  single_working_buoy <- sepmonthwp %>%
                         filter( PLATFORM_NUMBER == usefulBuoys$PLATFORM_NUMBER[i] &
                                 LONGITUDE == usefulBuoys$LONGITUDE[i] &
                                 LATITUDE == usefulBuoys$LATITUDE[i] )
  
  dist_to_thermocline <- abs( single_working_buoy$TEMP_ADJUSTED - usefulBuoys$TEMP_THERMOCLINE[i] )
  temp_sorting_value <- sort( dist_to_thermocline )[3]
  
  rows_near_therm <- single_working_buoy[ dist_to_thermocline < temp_sorting_value , ]
  
  usefulBuoys$THERMOCLINE_DEPTH[i] <- min( rows_near_therm$PRES_ADJUSTED )

}

truedepth <- usefulBuoys$THERMOCLINE_DEPTH

preddepth <- sepmonthwp %>%  group_by(PLATFORM_NUMBER , LONGITUDE , LATITUDE) %>%
               summarise(  thermpred= mean(thermpred) ) 
final <- inner_join(usefulBuoys,preddepth,by=c("LONGITUDE","LATITUDE","PLATFORM_NUMBER"))

final


```



What we need to do for model 2



BIN OUR AREA into independent bins essentially (this should be justified somehow)

Take all the typhoon data, and count how many on average each bin gets hit with a typhoon.

Now get the average thermocline depth in each bin. 

Now regress the count of typhoon this can't be binomial since we don't really have "trials" 
per say, maybe we do??? Regress the average count of typhoons on the thermocline depth. Observe if there is relationship. In essence this allows us to see the relationship but 
I don't think it lets us see the probability and of course this isn't causal. We would need to look at other confounders if we want that.

